Poincaré and SimBio: a versatile and extensible Python ecosystem for modeling systems.

### 2024-04-03 article.tex
Add compilation time
Change hardware to Mac M2 pro

To evaluate SimBio's performance,
we selected \ac{SBML} models from the curated section of BioModels \cite{malik-sheriffBioModels15Years2020}.
Among the first 250 models,
we considered the 117 that used supported \ac{SBML} features.
We ran simulations
on a Intel Xeon CPU E5-2620 v4
using
COPASI v4.42.284
(with BasiCO v0.64) with the LSODA solver,
RoadRunner v2.5.0 (with Tellurium v2.2.10) with the comparable CVODE solver, and
SimBio v0.3.2 with the LSODA solver.
For SimBio, we considered three variants:
NumPy (v1.26.4) backend and scipy (v1.12)'s LSODA solver,
Numba (v0.59.0) backend and scipy's LSODA solver,
Numba backend and numbalsoda's LSODA solver.
In all cases, we used a relative tolerance of \(10^{-6}\) and absolute tolerances of \(10^{-9}\).
We measured two simulation stages:
an initial cold run that includes the reading of the \ac{SBML} model
and subsequent warm runs.

For COPASI and Tellurium,
we noted that its runtime depended on the number of intermediate evaluation points (Figure~\ref{fig-runtime}, left).
For SimBio,
its NumPy backend can be orders of magnitude slower than both COPASI and RoadRunner  (Figure~\ref{fig-runtime}, right).
Nevertheless, switching to the numba backend,
which \ac{JIT} compiles the \ac{RHS} equations,
puts it on par with them.
Another speed-up in the runtime can be had by switching the LSODA \texttt{scipy} solver
for a more efficient \texttt{numbalsoda} implementation that avoids calling into the Python interpreter
between each of the integration steps.
A user might have to consider the trade-off between compilation and run times,
as the compilation of the \ac{RHS} code might take longer than the runtime itself,
and not be worth it for running the model only once.

>>>>

To evaluate SimBio's performance,
we selected \ac{SBML} models from the curated section of BioModels \cite{malik-sheriffBioModels15Years2020}.
Among the first 250 models,
we considered the 117 that used supported \ac{SBML} features.
We ran simulations
on a MacBook Pro with M2 Pro CPU
using
COPASI v4.42.284
(with BasiCO v0.64) with the LSODA solver,
RoadRunner v2.5.0 (with Tellurium v2.2.10) with the comparable CVODE solver, and
SimBio v0.3.2 with the LSODA solver.
For SimBio, we considered three variants:
NumPy (v1.26.4) backend and scipy (v1.12)'s LSODA solver,
Numba (v0.59.0) backend and scipy's LSODA solver,
Numba backend and numbalsoda's LSODA solver.
In all cases, we used a relative tolerance of \(10^{-6}\) and absolute tolerances of \(10^{-9}\).
We measured two simulation stages:
an initial cold run that includes the reading of the \ac{SBML} model
and subsequent warm runs.

For COPASI and Tellurium,
we noted that its runtime depended on the number of intermediate evaluation points (Figure~\ref{fig-runtime}, left).
For SimBio,
its NumPy backend can be orders of magnitude slower than both COPASI and RoadRunner  (Figure~\ref{fig-runtime}, right).
Nevertheless, switching to the numba backend,
which \ac{JIT} compiles the \ac{RHS} equations,
puts it on par with them.
Another speed-up in the runtime can be had by switching the LSODA \texttt{scipy} solver
for a more efficient \texttt{numbalsoda} implementation that avoids calling into the Python interpreter
between each of the integration steps.
A user might have to consider the trade-off between compilation and run times,
as the compilation of the \ac{RHS} code might take longer than the runtime itself,
and not be worth it for running the model only once.
For the models considered,
amortizing the compilation time required from 2 upto 200 runs.
